Artificial Ecosystem Selection Update

### 2020-10-21 Presentation/presentation.Rmd
update presentation

- Show you what I've been working on
- Hear your questions to see if that guides me down any path
- ???

# The Question

## 
One of the fundamental questions in ecology is, "How does the biodiversity
  that we observe influence ecosystem functions?"

## 
Historically, attempts to address this question have focused on alpha
  diversity as it relates to the *rate* or *stability* of an ecosystem function

##
For certain functions, this works pretty well. For example, the stability of
  primary productivity in temperate plant communities in the face of drought is
  correlated with species richness of plants.

## 
However, microbial community data exhibit certain limitations that cause this
  approach to fail:

  1. Without a consistent species concept, it is not obvious at what level of
     biological organization to evaluate this relationship. (the ASV? the
     genus? the family?)
  2. It is not always clear which members of a community are members of a
     particular functional group. (This can be determiend morpholigically for
     plants, but must be determined genetically in natural microbial
     communities.)
  3. Microbial communities are immensely diverse and without solutions to the
     previous two problems, sifting through that diversity is challenging.

## When we apply this BEF approach to microbial communities, there are limited
  gains in terms of our knowlegde of the controls on ecosystem function

# A new approach to linking microbial communities to ecosystem function

## An alternative approach

- One way to think about microbial communities and ecosystem functions is as assemblages of genes within an ecosystem and a trait at the ecosystem level.
- In fact, this often the only way we can measure microbial communities in natural ecosystems.

## Experiment

I collected a single soil core from near UO campus and used sterile potting mix as the subtrate. Ecosystems were created in 500 mL mason jars with 45 g of sterile potting mix and 5 g of living soil and were brought to 60% of field capacity (the optimal level for aerobic processes such as methane oxidation). I created 24 jars in this way and randomly assigned them to a Positive or Neutral selection line.

@@

fluxes <- read_tsv('../Output/fluxes.tsv')
ggplot(fluxes, aes(x = passage, y = estimate, color = treat)) + 
  geom_jitter() +
  scale_color_manual(name = 'Treatment', labels = c('Neutral', 'Positive'),
                     values = c('gray60', 'darkorange2')) +
  labs(x = "Passage", y = 'Flux (-k)') +
  theme_bw()

@@

fluxes %>% 
  select(treat, passage, estimate) %>% 
  group_by(passage, treat) %>% 
  summarize(mean = mean(estimate),
            se = se(estimate)) %>% 
ggplot(mapping = aes(x = passage, y = mean, ymin = mean - se, ymax = mean + se,
                     color = treat)) + 
  geom_pointrange() +
  scale_color_manual(name = 'Treatment', values = c('gray60', 'darkorange2')) +
  theme_bw()
```

@@

There was a response to selection on soil methane oxidation rate in the positive treatment relative to the neutral treatment.

@@

ggplot(deviance, aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_ribbon(aes(ymin = dev_pred$ci.lb, ymax = dev_pred$ci.ub), alpha = 0.2) +
  geom_pointrange(color = 'darkorange2') + 
      geom_segment(aes(x = 1, xend = passages, y = dev_fit$beta[1] + dev_fit$beta[2], yend = dev_fit$beta[1] + dev_fit$beta[2]*passages), color = 'darkorange2', size = 1) +
  labs(x = "Passage Number", y = "Deviation from Control (P - N)")+
  theme_bw()

```{r}
init <- 
  fluxes %>% 
  group_by(passage) %>% 
  summarize(mean = mean(estimate)) %>% 
  filter(passage == 1) %>% 
  pull()

@@

(initial mean flux  = `r myround(init, 3)` $\pm$ `r myround(init_se, 3)`)

>>>>

- Give the rationale/motivation for this project, what I've learned so far, and what I could learn from future data.
- Hear  your questions and comments that might guide future steps.
- Talk about what "heritability" means in a community context.

# How does biodiversity influence ecosystem function?

## One approach
Relating alpha diversity within a functional group to the *rate* or *stability* of an ecosystem function

```{r, fig.width=4, fig.height=3, out.width=6e2, out.height=4.5e2}
knitr::include_graphics('draught-stability.png')
```

## Reasons this works

1. There's a clear unit of biological organization - the species. 
2. It's obvious what a "plant" is.
3. Communities are relatively simple - (1-25 species).

## Limitations of microbial community data

  1. Without a consistent species concept, it is not obvious at what level of
     biological organization to evaluate this relationship. (the ASV? the
     genus? the family?)
  2. It is not always clear which members of a community are members of a
     particular functional group. Further, in a metabolic network, it's not
     always obvious which functional group is limiting the end product.
  3. Microbial communities are immensely diverse (100s-1000s of ASVs) and
     without solutions to the
     previous two problems, sifting through that diversity is challenging.

## An alternative conceptualization of communities

A community is a pool of genes analogous to a genome (the metagenome) and that community exhibits a phenotype (ecosystem function).

Thinking about it this way, an ecosystem function might exhibit similar properties to an organismal trait.

$V_P = V_G + V_E$

## Heritability is "the relative importance of heredity in determining phenotypic values"

"A character can be 'hereditary' in the sense of being **determined by
the genotype**[1] or in the sense of being **transmitted from parents to offspring**[2]."

[1] Broad-sense heritability,  $H^2 = V_G / V_P$

[2] Narrow-sense heritability, $h^2 = V_A / V_P$

<small>Falconer and MacKay, 1996</small>

# An alternative approach to linking microbial communities to ecosystem function

## Artificial Ecosystem Selection

- Generate lots of communities with different combinations of taxa.
- Put them into a common environment to minimize environmental variation.
- Test which combinations generate the greatest function.
- Reshuffle the best communities and test again.
- Similar to dilution to extinction or enrichment culturing.

## Method

To do this, I generated 24 ecosystems.

```{r, fig.width=4, fig.height=3, out.width=9e2*0.8, out.height=4.5e2*0.8}
knitr::include_graphics('aes_design.png')
```

@@

fluxes <- read_tsv('../Output/fluxes.tsv')
ggplot(fluxes, aes(x = passage, y = estimate, color = treat)) + 
  geom_jitter() +
  scale_color_manual(name = 'Treatment', labels = c('Neutral', 'Positive'),
                     values = c('gray60', 'darkorange2')) +
  labs(x = "Passage", y = 'Methane Oxidation Rate (-k)') +
  theme_classic()

@@

fluxes %>% 
  select(treat, passage, estimate) %>% 
  group_by(passage, treat) %>% 
  summarize(mean = mean(estimate),
            se = se(estimate)) %>% 
ggplot(mapping = aes(x = passage, y = mean, ymin = mean - se, ymax = mean + se,
                     color = treat)) + 
  geom_pointrange() +
  scale_color_manual(name = 'Treatment', labels = c('Neutral', 'Positive'),
                     values = c('gray60', 'darkorange2')) +
  labs(x = "Passage", y = 'Methane Oxidation Rate (-k)') +
  theme_classic()
```

@@

There was a response to selection on soil methane oxidation rate in the positive treatment relative to the neutral treatment.
<style>
.container{
    display: flex;
}
.col{
    flex: 1;
}
</style>

<div class="container">

<div class="col">
```{r, fig.width=4, fig.height=3, out.width=6e2*0.8, out.height=4.5e2*0.8}

@@

ggplot(deviance, aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_ribbon(aes(ymin = dev_pred$ci.lb, ymax = dev_pred$ci.ub), alpha = 0.2) +
  geom_pointrange(color = 'darkorange2') + 
      geom_segment(aes(x = 1, xend = passages, y = dev_fit$beta[1] + dev_fit$beta[2], yend = dev_fit$beta[1] + dev_fit$beta[2]*passages), color = 'darkorange2', size = 1) +
  labs(x = "Passage Number", y = "Deviation from Control (P - N)")+
  theme_classic()

```
</div>

<div class="col">
```{r}
init <- 
  fluxes %>% 
  group_by(passage) %>% 
  summarize(mean = mean(estimate)) %>% 
  filter(passage == 1) %>% 
  pull()

@@

(initial mean flux  = `r myround(init, 3)` $\pm$ `r myround(init_se, 3)`)
</div>

</div>

# How *heritable* is soil methane oxidation rate as an ecosystem trait?

## Broad- vs. Narrow-sense heritability

$V_G/V_P$ is the extent to which phenotype is determined by the genotype.

$V_A/V_P$ is the extent to which phenotype is determined by the genes transmitted from the parents.

## Broad- vs. Narrow-sense heritability

$V_G/V_P$ is the extent to which phenotype is determined by the genotype.

**$V_A/V_P$ is the extent to which phenotype is determined by the genes transmitted from the parents.**

## Narrow-sense heritability

$h^2 = R / S$

$R$ = Response to selection, difference between parents and offspring

$S$ = Selection differential, difference between selected parents and all parents

```{r}
h2_per_gen <- read_tsv('../Output/h2_per_gen.tsv')

h2_per_gen_sum  <-
  h2_per_gen %>% 
  summarize(h2_med = median(h2), h2_mad = mad(h2))
```

```{r, fig.width=2.5, fig.height=2.5, out.width=3e2, out.height=3e2}
ggplot(h2_per_gen, aes(x = passage, y = h2)) +
  geom_point() +
  geom_pointrange(aes(x = 2.5, y = h2_med, 
                                  ymax = h2_med + h2_mad,
                                  ymin = h2_med - h2_mad),
  data = h2_per_gen_sum) +
  labs(x = 'Passage Number', y = expression(paste("Narrow-sense Heritability (",
                                                h^2, ")"))) +
  theme_classic()

Each small point is the $h^2$ for that passage and its offspring. The
larger central point is the median $h^2$ (+/- median absolute difference)
across all four passages. $h^2$ =  `r myround(h2_per_gen_sum$h2_med, 2)` $\pm$ `r myround(h2_per_gen_sum$h2_mad, 2)`.

# Conclusions and interpetation

## Community composition impacts the rate of this ecosystem function

```{r, fig.width=4, fig.height=3, out.width=6e2*.8, out.height=4.5e2*.8}

ggplot(deviance, aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_ribbon(aes(ymin = dev_pred$ci.lb, ymax = dev_pred$ci.ub), alpha = 0.2) +
  geom_pointrange(color = 'darkorange2') + 
      geom_segment(aes(x = 1, xend = passages, y = dev_fit$beta[1] + dev_fit$beta[2], yend = dev_fit$beta[1] + dev_fit$beta[2]*passages), color = 'darkorange2', size = 1) +
  labs(x = "Passage Number", y = "Deviation from Control (P - N)")+
  theme_classic()

## `r myround(h2_per_gen_sum$h2_med, 2)` of the variation in methane oxidation rate is attributable to the additive effect of taxa passed from the parents

The remaining `r 1 - as.numeric(myround(h2_per_gen_sum$h2_med, 2))` is due to
environmental variation not controlled for experimentally or with the neutral
treatment, interactions among taxa, ...?

Caveat: this doesn't account for cor

```{r, fig.width=2.5, fig.height=2.5, out.width=3e2, out.height=3e2}
ggplot(h2_per_gen, aes(x = passage, y = h2)) +
  geom_point() +
  geom_pointrange(aes(x = 2.5, y = h2_med, 
                                  ymax = h2_med + h2_mad,
                                  ymin = h2_med - h2_mad),
  data = h2_per_gen_sum) +
  labs(x = 'Passage Number', y = expression(paste("Narrow-sense Heritability (",
                                                h^2, ")"))) +
  theme_classic()

## Next: Which taxa changed as a response to selection?
